% LaTeX for Steve Jarvis' CS 591

% The story I'm aiming to tell:
%
% INTRODUCTION:
%
% When it comes to international conflict, what have we relied upon to get us (the
% world) this far? Why can't we just keep doing that?
%
% Since we need something new, what should that something be?
%
% Others have recognized that cyberspace is new and important, and the
% most significant effort has been that of the Tallinn Manual. Introduce, briefly,
% the Tallinn Manual.
%
% ISSUE:
%
% Introduce Stuxnet, and highlight some of the complications it introduces when
% trying to apply traditional rules of conflict (jus ad bellum, jus in bello). We
% need something more.
%
% Can the Tallinn Manual satisfy our need for guidance in cyber conflict? Go over
% the help we need from the Tallinn Manual regarding Stuxnet.
%
% DETERMINE:
%
% How well does the Tallinn Manual apply to the case of Stuxnet? On which issues
% did it provide sufficient guidance, and where did it leave us wanting?
%
% ANALYSIS:
%
% It seems the shortcomings were simply issues not agreed upon by the experts
% or not yet resolved, rather than egregious oversights. It is generally impressively
% thoughtful and maintains the spirit of traditional laws and treaties when
% applying them to the Internet.
%
% PROPOSED SOLUTION:
%
% So, the international community should agree to a treaty which adopts the
% Tallinn Manual as:
% 1) An agreed upon conduct for cyber conduct.
% 2) A starting point for ruling on future conflict.
% The UN Security Council can be entrusted with ruling on future conflicts, as it
% currently does with kinetic conflict, using the Tallinn Manual as a general
% guide.
%
% Rulings made by the Security Council on cyber conflict should be considered
% international law, much as precedents build upon the US Constitution and
% previous interpretations in the United States to become law.
%
% Conclusion:
%
% This will be an effective solution to dealing with cyber conflict, as it is
% tailored towards continual evolution, which is exactly what we should expect in
% the age of the Internet.


% 12pt required for project
\documentclass[12pt]{article}

% Now the actual document
\newcommand{\hwtitle}{Applicability and Adoption of the Tallinn Manual}

\author{Steve Jarvis} \title{ \hwtitle{} }
\date{\today}

\usepackage{outlines}

% Doc formatting, double space with 1.25 sides and 1'' top/bottom
\usepackage{setspace}
%\doublespacing
\usepackage[margin=1.25in,top=1in,bottom=1in]{geometry}
\usepackage{enumerate}

% Page header
\usepackage{fancyhdr} \pagestyle{fancy} \lhead{Steve Jarvis, CS 591}
\rhead{\hwtitle{}}

% Bibliography
\usepackage[utf8]{inputenc}
\usepackage{biblatex-chicago}
\addbibresource{outline.bib}

\begin{document}
\maketitle

% The body of the outline
\begin{outline}
  \1 Introduction.
    \2 International law currently has a limited means for dealing with criminal
    cyber action.
      \3 The United Nations (UN) Charter and Geneva Conventions deal, historically,
      with only kinetic warfare.
        \4 Cyber actions rarely have a direct or obvious
        kinetic counterpart with which to compare. Differing interpretations
        results in dissension
        among states, with no clear answers on how to classify a use of force or
        armed attack online, or even whether it's possible to launch an armed
        attack in cyber space.\footnote{\cite{cyberAttacksYale}, pages 431-435.}
        \4 Some states adopt a broad view of what constitutes illegal
        force while others have developed a much narrower
        classification.\footnote{\cite{tallinnManual}, page 17, para 3.}
        \4 With no consensus on what constitutes illegal cyber activity, there
        can be no agreement on suitable international laws or treaties.
    \2 The Tallinn Manual aims to fill the void of defined international law
    or agreements when it comes to cyber
    activity.\footnote{\cite{tallinnManual}, page 18, para 2.}
      \3 The Tallinn Manual was the effort of an international group of experts,
      upon encouragement from the North Atlantic Treaty Organization
      (NATO), to correlate existing international laws to
      cyberspace.\footnote{\cite{tallinnManual}, page 16, para 1}

  \1 Description of the problem. How do we deal with major international cyber
  issues?
    \2 Consider, specifically, the actions and consequences of the Stuxnet
    Worm. Determine which, if any, international law was broken, and if so, how
    severely.
      \3 Explain Stuxnet technically, briefly.
        \4 Stuxnet was an attack to sabotage the programmable logic controllers
        (PLCs) at the Iran nuclear enrichment facility and delay
        Iran's potential development of nuclear
        weapons.\footnote{\cite{symantecStuxnetOverview}}
        \4 Stuxnet was not only extremely technically complex, it was also
        extremely expensive.\footnote{\cite{stuxnetCostComplexity}} As such,
        it's believed to have been conducted by a nation-state.
      \3 Why Stuxnet deserves the focus.
        \4 Prime alternative operations include Flame Malware
        and Russia's cyber attack on
        Georgia.\footnote{\cite{flameForbes}}\footnote{\cite{russiaAttackGeorgiaAC}}
        Flame was used for only espionage, and that is generally considered
        legal under international
        law.\footnote{\cite{tallinnManual}, rule 6, comment 4.} Russia's cyber
        attack on Georgia, conversely, was used to further a kinetic attack and
        falls clearly into the category of armed
        conflict.\footnote{\cite{tallinnManual}, page 94, para 6 and
          \cite{tallinnManual}, rule 20.} These conclusions are unsurprising and
        do not feel controversial.
      \3 What were Stuxnet's goals and what it accomplished.
        \4 Stuxnet aimed to damage to and slow development of Iran's nuclear
        development.\footnote{\cite{stuxnetGoalResult}}
        \4 Iran refused to comment on whether it suffered damage from
        Stuxnet. It's widely held that their nuclear program
        did suffer significant setbacks.\footnote{\cite{stuxnetIsis}}
      \3 Legal issues related to Stuxnet, on which there exists little or no
      help from established international law and guidance is desired from the
      Tallinn Manual.
        \4 Intentional damage caused to a sovereign state's infrastructure by a
        computer virus. Was it illegal, a use of force, and/or an ``armed attack''?
        \4 The damage was intended to stifle development of military operations.
        How severe of a response (if any) is warranted?
        \4 The worm leaked into the wild and civilians, Kapersky even claimed
        it's infected Russian nuclear
        facilities.\footnote{\cite{stuxnetLeakKapersky}} What repercussions
        might this entail?
        \4 In kinetic warfare, belligerents cannot use neutral territory for
        transport of munitions, and it is a duty of the neutral state to ensure
        that is the case.\footnote{\cite{hagueConventionV5}, articles 2 and 5.}
        Is it legal to move malware through the infrastructure of neutral
        states, and can those states be held responsible for allowing it?

  \1 Determine whether the Tallinn Manual is an effective tool beyond a
  theoretical use by finding whether it retains usefulness when applied to
  Stuxnet.
    \2 Where the Tallinn Manual meets the needs of an international
    cyber conflict doctrine.
      % Remember, an armed attack exceeds a use of force.
      \3 Unanimously determined to be in violation of international law and a
      use of force by the group of
      experts.\footnote{\cite{stuxnetForceWired2013}}
      \3 The manual goes on to say that the experts were unanimous in
      determining cyber operations alone might qualify as armed
      conflict.\footnote{\cite{tallinnManual}, rule 22, comment 15.} Tallinn
      Manual states:
      \begin{quote}
        An armed conflict exists whenever there are here are hostilities, which
        may include or be limited to cyber operations, occurring between two or
        more States.\footnote{\cite{tallinnManual}, rule 22.}
      \end{quote}
      \3 Tallinn Manual also comments directly that the experts were divided
      on whether Stuxnet qualified as an
      ``armed attack''.\footnote{\cite{tallinnManual}, rule 22, comments 14
        and 15.} Some experts felt it was justified as anticipatory
      self-defense.\footnote{See rule 13, comment 13. \cite{tallinnManual}}
      \3 Indiscriminate methods of cyber war are
      prohibited.\footnote{\cite{tallinnManual}, rule 43.} The infection of
      civilians' computers does not violate this, because despite infection, the
      attack was performed only on specific Siemens
      systems (not civilian).\footnote{\cite{stuxnetGoalResult}}
        \4 If Kapersky's claims that Stuxnet later infected Russian nuclear
        sites are true, however, Stuxnet did violate this rule, since it proves
        it was not limited in its effects as required by the law of armed
        conflict.\footnote{\cite{stuxnetLeakKapersky} and
          \cite{genevaArticle51}}
    \2 Issues on which the Tallinn Manual was not conclusive.
      \3 What self-defense is legal when it's unclear whether an operation could
      be classified as an ``armed attack''?
      \3 How might we determine what constitutes an armed cyber attack?
        \4 Tallinn Manual is clear that cyber operations alone could amount to
        an armed attack,\footnote{\cite{tallinnManual}, rule 22.} but provide no
        means to determine whether a specific operation crosses that threshold,
        or how an armed attack relates to a use of force (though that is a
        standing problem with kinetic warfare, just the same.)
      \3 Who can be held responsible in a cyber attack?
        \4 Tallinn Manual doesn't seem to mention a requirement of cyber
        combatants to identify themselves, and it's seems unlikely they would,
        anyway. Attribution is a big challenge.
        \4 What of neutral states whose infrastructure transferred the malware?
        Laws on kinetic warfare imply they can be held accountable, but
        cyberwar feels different in this regard. The Hague Conventions
        itself limits a neutral state's obligation to restrict the combatant use
        of telephone or wireless cables.\footnote{See article 8.
          \cite{hagueConventionV5}} Tallinn Manual states it must not
        \emph{knowingly} allow conflicting parties to use its
        infrastructure.\footnote{\cite{tallinnManual}. Rule 93.} The experts
        were split on whether that permits neutral states to passively allow
        such transmission.\footnote{\cite{tallinnManual} Rule 93,
          comments 5 and 6.}

  \1 Proposed solution for resolving international cyber conflict.
    \2 The Tallinn Manual does an outstanding job of applying existing
    international law to the age of the internet, albeit lacking some
    finer detail. Much of the gray area is equally undecided in kinetic
    conflict, and without application to real life international conflict
    it isn't obvious the Manual could reasonably be made much more
    specific.
      \3 For example, there is little guidance on determining if Stuxnet,
      or any operation, qualifies as an armed attack. Some views, seemingly
      including that of the United States (U.S.),\footnote{\cite{cyberAttacksYale},
        page 433, para 4.} suggest that a use of force and an armed attack
      may be one in the same, even regarding kinetic
      conflict.\footnote{\cite{kohSpeech}, page 7, para 5}
        \4 Schmitt says these specifics were intentionally not included in the
        Tallinn Manual.\footnote{\cite{areThereIntlRules}}
        \4 If we can assume an armed attack accompanies any use of force and
        warrants a response under Article 51, then the Tallinn Manual provides a
        fairly complete template to determine whether an operation
        qualifies.\footnote{\cite{tallinnManual}, rule 11, esp. comment 9.}
        \4 Schmitt claimed that:
        \begin{quote}
          [...]the majority said that an attack, in the law of war, means
          you physically harm someone, you break something, you cause physical
          damage or you interfere in the functionality of an object such that it
          needs to be actually repaired.\footnote{\cite{areThereIntlRules},
              para 4.}
        \end{quote}
        This seems to qualify Stuxnet as an armed attack, yet other experts
        still disagreed it crossed the undefined threshold. %TODO why disagree??
    \2 The Tallinn Manual should be adopted as a starting point for ruling on
    international cyber conflict and used to evolve international law, just as
    the U.S. Constitution served as a starting point for U.S. law and was molded
    by hundreds of years of precedence.
    \2 The current members of the UN should be encouraged to sign a treaty
    adopting adherence to the Tallinn Manual.
    \2 Violation of the Tallinn Manual should be ruled on by the UN Security
    Council, at which point any retaliation, up to and including that under
    Article 51, is a potential outcome.
    \2 There should be an official means to make lasting and binding amendments
    to the Manual, as deemed necessary by the Security Council and members of
    the UN.

  \1 Analysis of the solution.
    \2 The solution is neither complete nor final, but as with all law, we must
    set up a solid base with which to work and progress over time. The
    established international law made preparation for the age of the internet
    possible, and in the same way, the Tallinn Manual sets the stage for the
    tools we need in decades to come, as the world grows increasingly more
    reliant on the Internet.
    \2 Schmitt implied he believes the standards for classifying armed attacks
    in cyberspace will evolve in coming years:
    \begin{quote}
      I anticipate that we'll see a lot of thresholds coming down that will
      allow states to respond more vibrantly to cyber attacks that  might not be
      possible under the law as we found it.\footnote{\cite{areThereIntlRules},
        para 23.}
    \end{quote}
    \2 Achieving concurrence on acceptable cyber acts will likely prove
    challenging, just as international treaty acceptance tends to be, but it is
    a necessary feature as the cyber space continues to evolve in ways not yet
    considered.

  \1 Conclusion
    \2 The Tallinn Manual needs refinement and official acceptance, but is a
    tool worthy of founding international law and standards.
    \2 The efforts of the Tallinn Manual will give us confidence in the
    international arena that we, collectively, are continuing to act in cyber
    space in the spirit of already established and accepted laws. It is the
    confidence to move forward.

\end{outline}

\newpage
\nocite{*}
\printbibliography

\end{document}
