% LaTeX for Steve Jarvis' CS 591

% The story I'm aiming to tell:
%
%% INTRO what are we doing?:
%
% Highlight some of the complications cyber attacks introduce when
% trying to apply traditional rules of conflict (jus ad bellum, jus in bello).
% We need something more.
%
% Introduce the Tallinn Manual. Can the Tallinn Manual satisfy our need for
% guidance in cyber conflict?
%
% Goal of this report is to decide if the Tallinn Manual is suitable for official
% adoption in international conflict.
%
%% EXPLANATION of the issue, what's Stuxnet and what's lacking in traditional law?:
%
% Examine Stuxnet details. Why is Stuxnet special?
%
% Go over the help we need from the Tallinn Manual regarding Stuxnet. Where does
% traditional law miss and what do we need from Tallinn Manual?
%
%% DETERMINE the applicability of Tallinn Manual to Stuxnet. Does it work?
%
% How well does the Tallinn Manual apply to the case of Stuxnet? On which issues
% did it provide sufficient guidance, and where did it leave us wanting?
%
% What can we decide with Tallinn's help, and does it seem right?
%
%% PROPOSED SOLUTION is it's effective, what should we do about it?:
%
% It seems the shortcomings were simply issues not agreed upon by the experts
% or not yet resolved, rather than egregious oversights. It is generally impressively
% thoughtful and maintains the spirit of traditional laws and treaties when
% applying them to the Internet.
%
% So, the international community should agree to a treaty which adopts the
% Tallinn Manual as:
% 1) An agreed upon conduct for cyber conduct.
% 2) A starting point for ruling on future conflict.
% The UN Security Council can be entrusted with ruling on future conflicts, as it
% currently does with kinetic conflict, using the Tallinn Manual as a general
% guide.
%
% Rulings made by the Security Council on cyber conflict should be considered
% international law, much as precedents build upon the US Constitution and
% previous interpretations in the United States to become law.
%
%% ANALYSIS:
%
% Sounds like a good start.
%
%% CONCLUSION:
%
% This will be an effective solution to dealing with cyber conflict, as it is
% tailored towards continual evolution, which is exactly what we should expect in
% the age of the Internet.
%


% 12pt required for project
\documentclass[12pt]{scrreprt}

% Now the actual document
\newcommand{\hwtitle}{Applicability and Adoption of the Tallinn Manual}
\newcommand{\hwsubtitle}{Considering Legality and Repercussions of the Stuxnet Worm}

\author{Steve Jarvis}
\title{ \hwtitle{} }
\subtitle{ \hwsubtitle{} }
\date{\today}

\usepackage{outlines}

% Doc formatting, double space with 1.25 sides and 1'' top/bottom
%% See if any of this is actually necessary in final doc or if it just does it.
%\usepackage{setspace}
%\doublespacing
%\usepackage[margin=1.25in,top=1in,bottom=1in]{geometry}
\usepackage{enumerate}

% Page header
\usepackage{fancyhdr} \pagestyle{fancy} \lhead{Steve Jarvis} \rhead{\hwtitle{}}

% Bibliography
\usepackage[utf8]{inputenc}
\usepackage{biblatex-chicago}
\addbibresource{outline.bib}

\begin{document}
\maketitle

% The body of the outline
\begin{outline}
  \1 Introduction.
    \2 How can we effectively deal with the new world of international cyber
    operations?
    \2 International law currently has a limited means for dealing with criminal
    cyber action.
      \3 The United Nations (UN) Charter and Geneva Conventions deal, historically,
      with only kinetic warfare.
        \4 Cyber actions rarely have a direct or obvious
        kinetic counterpart with which to compare. Differing interpretations
        results in dissension
        among states, with no clear answers on how to classify a use of force or
        armed attack online, or even whether it's possible to launch an armed
        attack in cyber space.\footnote{\cite{cyberAttacksYale}, pages 431-435.}
        \4 Some states adopt a broad view of what constitutes illegal
        force while others have developed a much narrower
        classification.\footnote{\cite{tallinnManual}, page 17, para 3.}
        \4 With no consensus on what constitutes permissible cyber activity, there
        can be no agreement on suitable international laws or treaties.
    \2 The Tallinn Manual aims to fill the void of defined international law
    or agreements when it comes to cyber
    activity.\footnote{\cite{tallinnManual}, page 18, para 2.}
      \3 The Tallinn Manual was the effort of an international group of experts,
      upon encouragement from the North Atlantic Treaty Organization
      (NATO), to correlate existing international laws to
      cyberspace.\footnote{\cite{tallinnManual}, page 16, para 1}
    \2 The goal of this report is to determine whether the Tallinn Manual is
    suitable for official adoption in the international community by examining
    its applicability to the Stuxnet worm.

  % Explanation of the issue, what's Stuxnet and what's lacking in traditional law?
  \1 Describe the issue. Explain Stuxnet, why it is the chosen event for
  consideration, and where guidance is needed from a cyber operations guide
  such as the Tallinn Manual.
    \2 Consider the actions and consequences of the Stuxnet
    Worm. Determine which, if any, international law was broken, and if so, how
    severely.
      \3 Explain Stuxnet technically, briefly.
        \4 Stuxnet was an attack to sabotage the programmable logic controllers
        (PLCs) at the Iran nuclear enrichment facility and delay
        Iran's potential development of nuclear
        weapons.\footnote{\cite{symantecStuxnetOverview}}
        \4 Stuxnet was not only extremely technically complex, it was also
        extremely expensive.\footnote{\cite{stuxnetCostComplexity}} As such,
        it's believed to have been conducted by a nation-state.
      \3 Why Stuxnet deserves the focus.
        \4 Stuxnet offers a more complex situation and is the most likely
        candidate for a cyber-only operation to be considered a use of force.
        \4 Prime alternative operations include Flame
        Malware\footnote{\cite{flameForbes}} and Russia's cyber attack on
        Georgia.\footnote{\cite{russiaAttackGeorgiaAC}}
        Flame was used for only espionage, and that is generally considered
        legal under international
        law.\footnote{\cite{tallinnManual}, rule 6, comment 4.} Russia's cyber
        attack on Georgia, conversely, was used to further a kinetic attack and
        falls clearly into the category of armed
        conflict.\footnote{\cite{tallinnManual}, page 94, para 6 and
          \cite{tallinnManual}, rule 20.} These conclusions are unsurprising and
        do not feel controversial.
      \3 What were Stuxnet's goals and what it accomplished.
        \4 Stuxnet aimed to damage to and slow development of Iran's nuclear
        development.\footnote{\cite{stuxnetGoalResult}}
        \4 Iran refused to comment on whether it suffered damage from
        Stuxnet. It's widely held that their nuclear program
        did suffer significant setbacks, on the order of what could have been
        accomplished with a kinetic attack.\footnote{\cite{stuxnetIsis}}
      \3 Legal issues related to Stuxnet, on which there exists little or no
      help from established international law and guidance is desired from the
      Tallinn Manual.
        \4 Intentional damage caused to a sovereign state's infrastructure by a
        computer virus. Was it:
        \begin{enumerate}
          \item Illegal?
          \item A use of force?\footnote{\cite{unCharter}. See Article 2(4).}
          \item An armed attack?\footnote{\cite{unCharter}. See Article 51.}
        \end{enumerate}
        \4 The damage was intended to stifle development of military operations.
        How severe of a response (if any) is warranted?
        \4 The worm leaked into the wild and infected civilian devices, Kapersky
        even claimed it's infected Russian nuclear
        facilities.\footnote{\cite{stuxnetLeakKapersky}} What repercussions
        might this entail?
        \4 In kinetic warfare, belligerents cannot use neutral territory for
        transport of munitions, and it is a duty of the neutral state to ensure
        that is the case.\footnote{\cite{hagueConventionV5}. See Articles 2 and 5.}
        Is it legal to move malware through the infrastructure of neutral
        states, and can those states be held responsible for allowing it?

  % Determine the applicability of Tallinn Manual to Stuxnet. Does it work?
  \1 Determine whether the Tallinn Manual is an effective tool in practice
  by finding whether it proves useful when evaluating Stuxnet.
    \2 Where the Tallinn Manual meets the needs of an international
    cyber conflict doctrine.
      % Remember, an armed attack exceeds a use of force.
      \3 Unanimously determined to be in violation of international law and a
      use of force by the group of
      experts.\footnote{\cite{stuxnetForceWired2013}} Stuxnet directly violates at
      least Rule 1 of the Tallinn Manual.\footnote{\cite{tallinnManual}. See
        Rule 1.}
      \3 The manual goes on to say that the experts were unanimous in
      determining cyber operations alone might qualify as armed
      conflict.\footnote{\cite{tallinnManual}. See Rule 22, comment 15.} Tallinn
      Manual states:
      \begin{quote}
        An armed conflict exists whenever there are here are hostilities, which
        may include or be limited to cyber operations, occurring between two or
        more States.\footnote{\cite{tallinnManual}. See Rule 22.}
      \end{quote}
      \3 Tallinn Manual also comments directly that the experts were divided
      on whether Stuxnet qualified as an
      ``armed attack''.\footnote{\cite{tallinnManual}. See Rule 22, comments 14
        and 15.} Some experts felt it was justified as anticipatory
      self-defense.\footnote{\cite{tallinnManual} See Rule 13, comment 13.}
      \3 Indiscriminate methods of cyber war are
      prohibited.\footnote{\cite{tallinnManual}. See Rule 43.} The infection of
      civilians' computers (might\footnote{\cite{tallinnManual}. See Rule 1,
        comment 6.}) not violate this, because despite infection, the
      attack was performed only on specific Siemens
      systems (not civilian).\footnote{\cite{stuxnetGoalResult}}
        \4 If Kapersky's claims\footnote{\cite{stuxnetLeakKapersky}} that
        Stuxnet later infected Russian nuclear sites are true, however, Stuxnet
        did violate this rule, since Stuxnet was decidedly a cyber attack and
        this would prove it was not limited in its effects as required by the
        law of armed conflict.\footnote{\cite{genevaArticle51}}
    \2 Issues on which the Tallinn Manual was not conclusive.
      \3 What self-defense is legal when it's unclear whether an operation could
      be classified as an ``armed attack'', or when the Security Council has not
      yet reached an agreement?
      \3 How might we determine what constitutes an armed cyber attack?
        \4 Tallinn Manual is clear that cyber operations alone could amount to
        an armed attack,\footnote{\cite{tallinnManual}, rule 22.} but provide little
        means to determine whether a specific operation crosses that threshold,
        or how an armed attack relates to a use of force (though that is a
        standing problem with kinetic warfare, just the same). Schmitt says
        these specifics were intentionally not included in the
        Tallinn Manual.\footnote{\cite{areThereIntlRules}}
        \4 If we can assume an armed attack accompanies any use of force and
        warrants a response under Article 51 (similar to the apparent view of
        the United States (US)\footnote{\cite{cyberAttacksYale}.
        See page 433, para 4.}), then the Tallinn Manual provides a
        fairly complete template to determine whether an operation
        qualifies.\footnote{\cite{tallinnManual}. See rule 11, esp. comment 9.}
        \4 Schmitt claimed that:
        \begin{quote}
          [...]the majority said that an attack, in the law of war, means
          you physically harm someone, you break something, you cause physical
          damage or you interfere in the functionality of an object such that it
          needs to be actually repaired.\footnote{\cite{areThereIntlRules},
              para 4.}
        \end{quote}
        This seems to qualify Stuxnet as an armed attack, yet other experts
        still disagreed it crossed the undefined threshold. There was little
        physical damage done, compared traditional, kinetic
        force.\footnote{\cite{stuxnetArmedAttackRegister}}
        \4 There are also issues with immediacy that were not matters in kinetic
        warfare. Iran didn't know it was being attacked until the threat was
        over, and with no imminent or current attack there is no justification
        for an armed response.\footnote{\cite{tallinnManual}. See rule 15.}
      \3 Who can be held responsible in a cyber attack?
        \4 Iran obviously had jurisdiction in Stuxnet.\footnote{\cite{
            tallinnManual}. See Rule 2} Attribution is still a significant
        challenge though.
        \4 What of neutral states whose infrastructure was used to transfer
        Stuxnet? Laws on kinetic warfare imply they can be held accountable, but
        cyberwar feels different in this regard. The Hague Conventions
        itself limits a neutral state's obligation to restrict the combatant use
        of telephone or wireless cables.\footnote{\cite{hagueConventionV5}.
          See article 8.} Tallinn Manual states it must not
        \emph{knowingly} allow conflicting parties to use its
        infrastructure.\footnote{\cite{tallinnManual}. See rule 93.} The experts
        were split on whether that permits neutral states to passively allow
        such transmission.\footnote{\cite{tallinnManual} See rule 93,
          comments 5 and 6.}
    \2 The conclusions of the Tallinn Manual feel reasonable.
      \3 Many of the issues not directly resolved are complex and likely require
      analysis on a case-by-case basis. The issues that were directly resolved
      are not only founded on accepted international law, but feel just when
      applied to the cyber world.

  % Proposed solution. So is it's effective, what should we do about it?
  \1 Proposed solution for resolving international cyber conflict.
    \2 The Tallinn Manual does an outstanding job of applying existing
    international law to the age of the internet, albeit lacking some
    detail. Much of the gray area is equally undecided in kinetic
    conflict, terribly complex and likely to evolve at a rapid pace.
    \2 The Tallinn Manual should be adopted as a starting point for ruling on
    international cyber conflict and used to evolve international law, just as
    the US Constitution served as a starting point for US law and was molded
    by hundreds of years of precedence.
      \3 Some of the currently conflicting issues (responsibility of neutral
      states, whether installing malware is itself a violation of sovereignty,
      etc) on which the experts were divided could be solidified with time and
      applied as general rules.
    \2 The current members of the UN should be encouraged to sign a treaty
    adopting adherence to the Tallinn Manual.
    \2 Violation of the Tallinn Manual should be ruled on by the UN Security
    Council, at which point any retaliation, up to and including that under
    Article 51, is a potential outcome. This is also suggested in the Tallinn
    Manual itself.\footnote{\cite{tallinnManual}, rule 18.}
    \2 There should be an official means to make lasting and binding amendments
    to the Manual (or treaty), as deemed necessary by the Security Council and
    members of the UN.

  \1 Analysis of the solution.
    \2 The solution is neither complete nor final, but as with all law, we must
    set up a solid base with which to work and progress over time. The
    established international law made preparation for the age of the internet
    possible, and in the same way, the Tallinn Manual sets the stage for the
    tools we need in decades to come, as the world grows increasingly more
    reliant on the Internet.
    \2 Schmitt implied he believes the standards for classifying armed attacks
    in cyberspace will evolve in coming years:
    \begin{quote}
      I anticipate that we'll see a lot of thresholds coming down that will
      allow states to respond more vibrantly to cyber attacks that  might not be
      possible under the law as we found it.\footnote{\cite{areThereIntlRules},
        para 23.}
    \end{quote}
    \2 Achieving concurrence on acceptable cyber acts will likely prove
    challenging, but it is a necessary feature as the cyber space continues to
    evolve in ways not yet considered.

  \1 Conclusion
    \2 The Tallinn Manual is worthy of official acceptance in founding
    international law and standards.
    \2 The international community can look forward to a healthy, global
    internet with help from the Tallinn Manual and subsequent discussions on
    permissible and responsible online operations and stewardship.
    \2 The efforts of the Tallinn Manual will give us confidence in the
    international arena that we, collectively, are continuing to act in cyber
    space in the spirit of already established and accepted international laws.
    It is the confidence to move forward.

\end{outline}

\newpage
\nocite{*}
\printbibliography

\end{document}
