% LaTeX for Steve Jarvis' CS 591

% 12pt required for project
\documentclass[12pt]{article}

% Now the actual document
\newcommand{\hwtitle}{International Adoption of the Tallinn Manual}
\newcommand{\hwsubtitle}{Examining Legality and Repercussions of the Stuxnet Worm}
\newcommand{\hwcompletetitle}{\hwtitle{} : \newline\hwsubtitle{}}

% To modify title spacing later
\usepackage{titling}

\author{Steve Jarvis}
\title{\hwcompletetitle{}}
\date{Boston University, CS 591: Cybersecurity and Policy \endgraf \today}

% Doc formatting, double space with 1.25 sides and 1'' top/bottom
\usepackage{setspace}
\usepackage[margin=1.25in,top=1in,bottom=1in]{geometry}
\usepackage{enumitem}
\setlist{nosep}

% Page header
\usepackage{fancyhdr} \pagestyle{fancy} \lhead{Jarvis} \rhead{\hwtitle{}}

% Bibliography
\usepackage[utf8]{inputenc}
\usepackage[notes]{biblatex-chicago}
\addbibresource{adopting-tallinn.bib}

% Say bibliography, not references
\DefineBibliographyStrings{english}{%
  references = {Bibliography},
}

\begin{document}

% Title a third of the way down. Name, class, date a few lines later.
\posttitle{\par\end{center}\vskip 2in}
\setlength{\droptitle}{1in} % Not absolute, just moves it by 'x' amount
\maketitle
\newpage

\tableofcontents
\newpage

% Double space the body
\doublespacing

% body
\section{Introduction}
Recent years have witnessed a shift of the earth's battlefield. For better or
worse, militaries can largely trade in their guns for laptops and their soldiers
for hackers. The nations of the world are connecting critical services and
infrastructure to cyberspace, and doing so with minimal technical or legal
protections in place. This puts them at great risk of attack. As the
offensive capabilities and risks continually increase in cyberspace, so
too must the protections.

In the wake of the Second World War, the international community recognized the
necessity of establishing agreements and rules governing international conflict.
The United Nations was organized in 1942
and the Geneva Conventions in
1949.\footnote{\cite{unitedNationsHistory}; \cite{genevaConventionsOrigins}}
These organizations founded treaties that define international laws for entering
into conflict (\textit{jus ad bellum}) and behavior during conflict
(\textit{jus in bello}), respectively. To this day, those laws protect
States from unjust attack and offer aid and support when the laws
are broken and a nation is victimized.

The agreements by these groups were established a full half-century before the
public Internet was in its infancy, and therefore the laws consider only kinetic
warfare.\footnote{``Kinetic'' is used throughout to describe traditional
  military action, one that involves physical movement and attack.} The
growing prevalence of cyber operations is pushing the
boundaries of what might be considered harmful or forceful behavior, and the
international community has little means to deal with such events. Cyber
operations rarely have a direct or obvious kinetic
counterpart for comparison, and without established guidelines or limitations,
States continue to progress their offensive cyber arms race. Cyber operations are
limited only by technical capabilities rather than legal or ethical standards.

The Tallinn Manual was created to fill this void in international law
by translating the established kinetic laws to this modern, cyber
battlefield.\footnote{\cite{tallinnManual}, 18. See paragraph 2.} The need
for such a set of laws grows more apparent with each passing year, as defensive
capabilities remain relatively stagnant while offensive arsenals and dependence
on cyberspace increase dramatically.

The goal of this report is to examine whether the Tallinn Manual is a
suitable guide for use in establishing agreements on permissible cyberspace operations
in the international community. To make this
determination, its applicability to real life actions, particularly the Stuxnet
worm, will be explored, as well as the Tallinn Manual's political hurdles. If
successful, the Tallinn Manual could be the basis of an official adoption of
cyberspace laws and help to ensure the
safety and security of all States and their citizens in coming years.

\section{The Current International Law}
Today, 193 of 196 countries of the world are members of the United
Nations (Kosovo, Taiwan, and Vatican City are the only non members).\footnote{\cite{unMembers}}
These countries enjoy agreements protecting their sovereignty and freedom from
unsolicited acts of physical aggression. These laws do not explicitly exclude
cyber acts of aggression, but the nature of international law is permissive; if
an act isn't explicitly stated illegal, then it is considered
legal.\footnote{\cite{lotusSummary}} Since there are no laws
established for cyber operations and little or no clear correlation to kinetic
laws exists, there is no obvious illegal cyber operation between nations.

States have developed differing opinions on how established laws effect cyber
operations. The different interpretations result in dissension among States and
incompatible policies. This lack of cohesion has serious results: there are no
accepted answers on whether it is possible to exert a use of force or launch an
armed attack online, how to qualify such attacks if they are a possibility, or
what repercussions would be appropriate if those attacks were
executed.\footnote{\cite{unCharter}; \cite{cyberAttacksYale}, 431-435. See the
  United Nations charter for definitions of a use of force and armed attack,
  specifically Articles 2(4) and 51.}

The United Nations Security Council rules on kinetic uses of force involving
members of the United Nations. The same cannot be afforded to nations victim of
a cyber attack, since there are effectively no relevant laws. There can be no
legitimate and organized repercussion from the international community.

\section{Why The International Community Needs to Establish Cyber Agreements} \label{need}
In 2007, Estonia chose to relocate a bronze statue of a Soviet soldier to
quell a dispute between Estonian nationalists, who saw the statue as a symbol
of oppression, and a Russian ethnic group, who considered the statue's take down
offensive. Following the statue's move, Russia initiated perhaps the largest
distributed denial of service (DDOS) attack seen to date. Estonia is one of the
most cyber connected nations in the world, and the attack disabled banking, news
websites, the government's online services, telephony and credit card
verification systems. Estonia's communications and commerce were rendered
effectively useless for weeks.\footnote{\cite{clarkeCyberWar}, 12-16.}

The events in Estonia demonstrate the capabilities of even simple cyber attacks,
and the serious detrimental effects for States as connected as
Estonia prove the concern is significant.\footnote{\cite{clarkeCyberWar},
  13. In at least 2007, Estonia ranked ahead of even the United States with
  regards to Internet use for daily tasks.}

The world's situation is growing more treacherous as States increase their
reliance on cyberspace. Critical infrastructure, including power
grids, water treatment plants, financial institutions and health care equipment,
is coming online and reliant on the Internet for communication and proper
operation. Citizens and governments depend on these services to maintain a
healthy and safe standard of living. This infrastructure, and by extension the
cyberspace that enables its proper operation, has a direct effect on a State's
sovereignty and deserves the same protections the physical world is granted.
For this same reason, the United States' National Research Council agreed that
attacks which have a significant effect on a State's infrastructure should be
considered armed attacks, even without direct injury, violence, or
death.\footnote{\cite{nrcCyberAttack}, 254.}

The new realm of cyberspace provides a battleground States of the world are
just beginning to explore. New attacks are constantly developed and far outpace
States' abilities to defend against them. It is crucial that the international
community does what it can by establishing agreements protecting States' cyberspace,
just as it did protecting States from traditional attacks.

\section{Overview of the Tallinn Manual}
The Tallinn Manual was an effort to address this growing concern over the threats
posed via cyberspace. Many nations, including the United Kingdom, the United
States, Russia, and Canada, cite national cyber security as an utmost
priority.\footnote{\cite{tallinnManual}, 16-17} The Tallinn Manual's aim
was to fill the void in the international laws on conflict by determining the
correlation between cyber operations and kinetic counterparts. Ultimately,
the stated goal of the Tallinn Manual is to provide non-binding guidance on
which cyber operations are within the bounds of the
law.\footnote{\cite{tallinnManual}, 16. See paragraph 1.}

The North Atlantic Treaty Organization (NATO) encouraged and supported the
project. Michael Schmitt, a professor at the United States Naval War College and
editor of the Tallinn Manual, led an international group of legal experts from many
NATO States in discussion regarding the complications of cyber operations and
how to apply established laws to this new
dimension.\footnote{\cite{tallinnManual}, 17. The group of experts will be referred to
  throughout as simply the ``Experts''.} The result is a
set of twenty-five rules and many additional thoughts, issues and sticking points in the
form of commentary throughout the document. The rules cover such issues as
defining a violation of sovereignty and jurisdiction, categorizing and defining
a use of force and cyber (armed) attack, and special considerations for
civilians.\footnote{\cite{tallinnManual}. See rules 1, 2, 10-11, 30 and 32,
  respectively.} All of the conclusions reached by the Tallinn Manual are
inspired by established international law, often specifically the United Nations
Charter and the Geneva Conventions, and the reasoning is often thoroughly
assessed in the embedded commentary.

The three-year effort concluded in 2012 and continues to stand alone in its
thoroughness in considering a modern world and outpaced laws. While the self-declared
intentions were to construct a non-binding document, practice has proven the
Tallinn Manual unparalleled in its applicability and usefulness. Timothy Edgar,
the first Director of Privacy and Civil Liberties for the White House National
Security Staff, states that ``... in reality, if you want to determine the legality of
a cyber operation, you reach for the Tallinn Manual.''\footnote{\cite{edgarTallinn}}

To determine if the Tallinn Manual is technically useful in addition to being
legally sound, it will be examined in the context of a real life cyber
operation: the Stuxnet worm.

\section{Details of the Stuxnet Worm}
Stuxnet was a revolutionary cyber attack executed in 2010 that sabotaged the programmable
logic controllers (PLCs) at Iran's nuclear enrichment facility at Natanz. The
intent was to delay Iran's potential development of
nuclear weapons, and it did this by causing the centrifuges that separate
nuclear material to spin out of control and ultimately destroy
themselves.\footnote{\cite{symantecStuxnetOverview}; \cite{stuxnetGoalResult}}

Despite the destructive chaos actually happening in the system, the Stuxnet
malware continuously delivered positive reports to the operators, providing
assurance that all operations were going safely and as planned. The human
operators capable of intervention had no indication there was an issue, much
less one so dire. Stuxnet was the first operation consisting of only a computer
program to cause destruction and setback to a sovereign State's military; it was
a weapon made entirely of code.

To consider Stuxnet an act of aggression by another nation it first has to be
known that it was, in fact, the product of a State, and not the effort of a private group of
citizens or hacktivists. Very early on there was strong evidence pointing
towards a State backing. Stuxnet's level of technical difficulty was uncontested
by any cyber attack prior (and maybe since), and in addition to being very
complex, it was also extremely
expensive.\footnote{\cite{stuxnetCostComplexity}. The cost of Stuxnet is
  estimated to be around one million U.S. dollars.}
More recently however, solid claims have surfaced proving it was
executed by a nation state, particularly a collaboration between the United
States and Israel.\footnote{\cite{snowdenStuxnetAuthor}} Stuxnet is now widely
accepted to have been orchestrated by a nation state, which qualifies it as a
politically motivated operation and subject to international laws of
conflict.

The enrichment facility at Natanz was not connected to the Internet and relied
on a private, ``air-gapped'' network.\footnote{``Air gap'' is a term commonly
  used to describe a network that is physically separate from unsecured
  networks.} Despite these precautions taken against
potential cyber attack, the facility proved to still be
vulnerable. Stuxnet was likely introduced to the target
environment by an infected USB drive.\footnote{\cite{stuxnetMicroscope}, 8.} The
route to Natanz is unknown, but it is likely the virus crossed
through the Internet infrastructure of a number of neutral States.

The damage was intended to stifle development of military operations by damaging
Iran's ability to enrich uranium. The precise effects are unknown, as Iran
made no official comment on whether it suffered damage from Stuxnet. However, it is
widely held that their nuclear program did suffer significant setbacks, on the
order of what might have been accomplished with a kinetic
attack.\footnote{\cite{stuxnetIsis}}

As with many worms, and almost by their very design, Stuxnet proved difficult
to control. It leaked into the wild and infected an untold number of civilian
and State owned computers in more than a dozen
countries.\footnote{\cite{stuxnetMicroscope}, 15.} Eugene Kapersky,
founder of the Russian security company Kapersky Labs, claimed the virus even
infected Russian nuclear facilities. He is not alone in these claims
either, with as many as fifteen other industrial complexes around the world
believed to have been infected by Stuxnet.\footnote{\cite{stuxnetPlants}}

The distinction between civilian and nuclear or industrial facility
infection is significant, not only due to concerns
regarding civilian targeting, but because of the nature of Stuxnet itself.
Stuxnet was designed to recognize the hardware on which it was running, and if
it wasn't on specific Siemen's PLCs it would sit
dormant.\footnote{\cite{stuxnetPlants}}
Civilians with infected computers would not only be unaware, they would be at no
risk of danger whatsoever. The infected industrial facilities, however, were
likely to suffer grave damage, just as Iran's facility is believed to have experienced at
Natanz.\footnote{\cite{stuxnetIsis}}

\subsection{Why Stuxnet Deserves the Focus}
Stuxnet exists amidst many controversial cyber operations over the past decade,
and it still deserves much focus when considering cyber operations and
international law on armed conflict. It offers, arguably, the
most likely candidate of a cyber-only operation to be considered a use of force
or an armed attack. It exists on a level of complexity and intent all its own. Prime
alternative operations include the Flame Malware and Russia's cyber attack on
Georgia.\footnote{\cite{flameForbes}; \cite{russiaAttackGeorgiaAC}}
Flame was used for only espionage, and that is generally considered legal under
international law.\footnote{\cite{tallinnManual}. See rule 6, comment 4.} Russia's
cyber attack on Georgia in 2008, conversely, was used to further a kinetic
attack and falls clearly into the category of an armed
attack.\footnote{\cite{tallinnManual}, 94. See paragraph 6. Also see rule 20.}
These operations raise many interesting questions of their own, but the overall
conclusions are unsurprising and generally uncontroversial.

\section{Where Traditional Law is Not Enough}
Cyber operations like Stuxnet are unlike any with which the world has previously dealt.
The executions of these attacks raise many questions where guidance from a
suitable source is desired to determine which laws, if any, were broken, how
severely they were broken, and what response might be warranted. There are a
handful of significant issues raised by Stuxnet which warrant specific
consideration:

\begin{enumerate}
  \item Was Stuxnet a use of force, as defined by international laws of conflict?
  \item If it was a use of force, does it also qualify as an armed attack?
  \item Did the operation lack sufficient discrimination and direction?
  \item Which States can be held accountable in such an event?
  \item If international laws were broken, what response is warranted by the
    victim State? How might the aggressor be reprimanded, and how should the
    rebuke be enforced?
\end{enumerate}

Each of these issues will be explored in depth with specific
consideration for what makes the issue controversial under the current laws of
conflict.

\subsection{Use of Force}
At perhaps the most fundamental, it is not clear whether Stuxnet was illegal under
international law. To determine its legality there would need to be consensus
that it did violate a law or laws already in place. One of the foremost
(though certainly not only) laws would be that prohibiting a use of force, set forth
by Article 2(4) of the United Nations Charter. Article 2(4) states:

\singlespacing
\begin{quote}
  All Members shall refrain in their international relations from the threat or
  use of force against the territorial integrity or political independence of
  any State, or in any other manner inconsistent with the Purposes of the United
  Nations.\footnote{\cite{unCharter}. See Article 2(4).}
\end{quote}
\doublespacing

Regarding only this law, which was established for dealing with kinetic
operations, Stuxnet's classification as a use of force is not
clear. Though there is no distinction between cyber and kinetic operations
mentioned, the traditional notion of force requires military force
and violence, whereas Stuxnet, and nearly all cyber attacks, did not even
include a physical presence, much less kinetic force. The
interpretations of laws as established are insufficient to categorize Stuxnet as
a use of force.

\subsection{Armed Attack}
Some States have adopted a generally all-encompassing definition of what constitutes
an illegal use of kinetic force, while others operate with much narrower
classifications that limit an armed attack to be effectively the same as a use
of force.\footnote{\cite{tallinnManual}, 17. See paragraph 3.}

One of the most popular definitions of an armed attack is resultant of a 1986
case decided by the International Court of Justice between the United States and
Nicaragua. The ruling classified an armed attack as one which is executed by
armed forces sent directly by a State or on behalf of a
State.\footnote{\cite{usVsNicaragua}. See paragraphs 187-201.}
This definition is rather limiting and seems to explicitly exclude any cyber attacks by its
very nature, since it makes numerous references to traditional arms. Similarly,
most all traditional definitions of armed conflict require some level of
physical violence matching or in excess of a use of
force.\footnote{\cite{nrcCyberAttack}, 253. See paragraph 2.} There have been no
cyber-only attacks to date that fit this description, and it very well may be
impossible simply by definition.

Furthermore, any conclusion to the issue of qualifying an armed attack is surely
more complex than a simple black and white answer. The infections of civilian machines could be
categorized differently than infections of State owned infrastructure. Civilian
infections were not necessarily harmful, and mere infection may not equate to a
cyber attack in any case.

Determining when an operation crosses the threshold and becomes an armed attack
is generally difficult, but as mentioned in section \ref{need}, it is crucial
the international community has the means to classify attacks in cyberspace.

\subsection{Discriminate Attack}
Supposing the Stuxnet worm can be qualified as an armed attack, its spread
throughout cyberspace raises questions of discrimination. A similar
failure to control weapons in kinetic warfare is outlawed as
indiscriminate, but it is not clear the same applies to
malware, since it often has far less directly devastating
results.\footnote{\cite{genevaArticle51}} With malware there is little danger of
immediate harm relative to indiscriminate kinetic attacks, such as acts of chemical
warfare or the destruction of a hospital. Even without immediate harm, though, if a
cyber operation can be classified as an attack, it is logical that it must
be subject to the same laws of discrimination.

\subsection{Accountability of Neutral States}
If an operation is decided to be in violation of any laws, the sponsors
of the malware and its authors should clearly be held responsible. There may be
additional participants to consider, though. In kinetic warfare,
belligerents cannot use neutral territory for transport of
munitions, and it is a duty of the neutral State to ensure that is the
case.\footnote{\cite{hagueConventionV5}. See Articles 2 and 5. The Hague
  Conventions are generally superseded by modern treaties, but are still
  cited as inspiration in many places of the Tallinn Manual.} Application of
those laws to cyberspace could require neutral States to take measures
preventing the transfer of malicious software or other facilitation of attacks
via cyber infrastructure located within their territory. It is possible that a
failure to stop the propagation of a cyber attack is, in itself, considered an
attack.

\subsection{Defense and Rebuke}
Equally important is the issue of enforcement and reprimanding States who are
deemed in violation of accepted law. The United Nations Charter allows for
a use of force as self defense in Article 51, though it also requires any
response to be proportional to the action which provoked
it.\footnote{\cite{unCharter}} No cyber operation
has been met with a military response to date, and in many cases such a
response would seem excessive. For clarity on
future policies, it is important to determine whether such a response could be
deemed reasonable.

International conflict between members of the United Nations is subject to
the United Nations Security Council for
review and ruling. On multiple occasions, however, the Security Council has
failed to reach a timely decision (or reach a decision at all), and this has
potentially devastating consequences for victim
States.\footnote{\cite{securityCouncilParalysis};
  \cite{securityCouncilDeadlockKosovo}. These are two examples of Security Council
  deadlock where no decision was reached.}
If a similar process is used for cyber operations it is important to consider
what actions a State can take, within the bounds of the law, to protect itself
in the time before an official ruling by the Security Council.

\section{Applying the Tallinn Manual to Analyze the Stuxnet Worm}
% Determine the applicability of Tallinn Manual to Stuxnet. Does it work?
Now we turn to the Tallinn Manual to resolve these issues applying existing
laws to cyberspace. The goal is to determine whether the Tallinn Manual is an
effective tool in practice by exploring whether it proves useful when evaluating
Stuxnet.

Regarding the legality of Stuxnet, the Tallinn Manual defines a cyber operation to
be a use of force when the
effects of the operation ``are comparable to non-cyber operations rising to the
level of a use of force.''\footnote{\cite{tallinnManual}, 48. See rule 11.} It goes on to suggest
a detailed rating scale for cyber operations in the comments, including such
factors as severity, immediacy, directness, invasiveness and military character,
most of which Stuxnet clearly violates.\footnote{\cite{tallinnManual},
  49-52. See comment 9.} The group of Experts considered Stuxnet specifically in
their work, and unanimously concluded it to
be a use of force. Therefore, the operation was in violation of international law by Article
2(4) of the United Nations Charter.\footnote{\cite{stuxnetForceWired2013}}

The manual goes on to say that the Experts were unanimous in determining cyber
operations alone might qualify as armed
conflict:

\singlespacing
\begin{quote}
  An armed conflict exists whenever there are hostilities, which
  may include or be limited to cyber operations, occurring between two or
  more States.\footnote{\cite{tallinnManual}, 71. See rule 22.}
\end{quote}
\doublespacing

This finding alone is significant, because it would bring responding to a cyber
attack with a military use of force within the realm of possibility. This
finding directly contradicts other interpretations of the Charter that reserve
an armed attack for kinetic military
operations.\footnote{\cite{internationalLawForCyber}. This is one such
  interpretation that reserves the category of armed attack for kinetic force.}

The case of Stuxnet is specifically mentioned in the commentary of the Tallinn
Manual regarding an armed attack. It is stated that the Experts were divided on
whether Stuxnet should be categorized as an armed attack, mostly due to an
insufficient level of damage caused.\footnote{\cite{tallinnManual}, 71. See
  rule 22, comments 14 and 15. There commentary also cited uncertainty that the
  operation was the attributable to a
  State and not an independent group, but this has since been generally
  confirmed.} Some Experts noted the attack may also have been justified as
anticipatory
self-defense.\footnote{\cite{tallinnManual}, 56. See rule 13, comment 13.}

The Tallinn Manual definitively decides that indiscriminate methods of cyber war are
prohibited, which is appropriate since it concluded that it is possible for cyber
operations to reach the level of an armed
attack.\footnote{\cite{tallinnManual}. See rule 43.} It is undecided whether
the mass infection of computers violates this, because the group of Experts was
undecided on ruling the mere installation of malware to be a violation of
sovereignty.\footnote{\cite{tallinnManual}, 25. See comment 6 of rule 1.}
In the case of Stuxnet, infection alone was unremarkable and the attack was
performed only on specific Siemens systems, so it is not clear whether even widespread
infection would put Stuxnet in violation.\footnote{\cite{stuxnetGoalResult}}

If Kapersky's claims that Stuxnet later infected Russian nuclear sites are true,
though, Stuxnet did violate the sovereignty of Russia and failed to be
sufficiently discriminate with its target.\footnote{\cite{stuxnetLeakKapersky}}

The Tallinn Manual is not equally conclusive on all matters, though. As
mentioned, Tallinn Manual is clear that cyber operations alone
could amount to an armed
attack, but provide little means to determine whether a specific operation
crosses that threshold, or how an armed attack relates to a use of
force.\footnote{\cite{tallinnManual}, 71. See rule 22.} These are standing problems
with kinetic conflict as well, and Schmitt says
these specifics were intentionally not expanded upon in the
Tallinn Manual, as to not alter the definitions provided by existing
international law.\footnote{\cite{areThereIntlRules}}

Despite offering little means for identifying a specific threshold, there is a
definition provided for a cyber (armed) attack. In an interview, Schmitt explained
that

\singlespacing
\begin{quote}
  ... the majority said that an attack, in the law of war, means
  you physically harm someone, you break something, you cause physical
  damage or you interfere in the functionality of an object such that it
  needs to be actually repaired.\footnote{\cite{areThereIntlRules}.
    See paragraph 4.}
\end{quote}
\doublespacing

The Tallinn Manual offers a similar definition, which is a

\singlespacing
\begin{quote}
... cyber operation, whether offensive or defensive, that is reasonably
expected to cause injury or death to persons or damage or destruction to
objects.\footnote{\cite{tallinnManual}. See rule 30.}
\end{quote}
\doublespacing

Both Schmitt's comment and the consensus of the legal scholars seem to suggest
Stuxnet was an armed attack, yet the Experts still disagree it crossed the
threshold of damage. Many believe there was little physical damage done,
compared to traditional, kinetic force.\footnote{\cite{stuxnetArmedAttackRegister}}

Iran had clear jurisdiction in
Stuxnet.\footnote{\cite{tallinnManual}, 27. See rule 2.} It is now widely accepted
that Stuxnet was the product of an effort between the United States and Israel,
and if Iran were seeking a response they would surely be the primary subjects.
Laws on kinetic warfare state that neutral States who fail to remain neutral can
also be held accountable for transferring munitions. Regarding the transfer of
malware, Tallinn Manual specifies
that States must not \emph{knowingly} allow conflicting parties to use their
infrastructure.\footnote{\cite{tallinnManual}, 206. See rule 93.} The Experts
were split on whether that permits neutral States to passively allow
such transmission. Many felt that in order to claim there was an unknown
transfer there must be a significant effort to ensure such activity is not
taking place.\footnote{\cite{tallinnManual}. See rule 93, particularly
  comments 5 and 6.} In the end, the Tallinn Manual provides no bottom
line on the responsibility of neutral states or their accountability for
passively transferring cyber munitions.

When considering a response, the issue of immediacy is more of a factor
in cyberspace than in kinetic warfare. In the case of Stuxnet, Iran didn't know
it was being attacked until the machinery was already damaged and the attack was
effectively complete. With no imminent or current attack there is
no justification for an armed response under Article
51. Conversely, a kinetic attack is generally immediately apparent to the
victim and immediacy is rarely a consideration. Again, the Tallinn Manual did
not change the definition of immediacy for cyberspace, only provided means to
help determine whether cyber operations meet the established standards for
immediacy.\footnote{\cite{tallinnManual}, 60. See rule 15, especially
  comments 9 and 10.} With the conclusions drawn by the Tallinn Manual, Iran
would only be able to respond with force if they believed the attack was part of
a larger campaign and likely to continue.\footnote{\cite{tallinnManual},
  62-63. See comment 9 of rule 15 on immediacy.}

\section{Why International Law Should Adopt the Tallinn Manual}
% Proposed solution. So is it is effective, what should we do about it?
While the Tallinn Manual was not conclusive across all issues, it does an
exceptional job of applying existing international law to the age of the
Internet. It provides a great foundation, but dealing effectively
with this era of cyber operations requires more than just translations of
laws from the kinetic era.

As an example of where more than a simple translation may be warranted, consider
again the 2007 conflict between Russia and Estonia. The Tallinn Manual is
unlikely to qualify that event as an armed attack,
since there was no reasonable expectation of injury, death, or
damage.\footnote{\cite{tallinnManual}. See rule 30.} As States grow more
dependent on cyberspace for the operation of critical infrastructure, however,
there is no longer a requirement of physical destruction or violence to threaten
a State's sovereignty. This was apparent in Estonia, as victims were unable to
access their own funds or purchase basic needs.

Therefore, this conclusion of
the Tallinn Manual is one that needs to be improved in order to catch up with
the age of the Internet. Michael Norris suggests an amendment to the Tallinn
Manual's definition of cyber attack to reconcile (difference emphasized):

\singlespacing
\begin{quote}
  A cyber attack is a cyber operation, whether offensive or defensive, that is
  reasonably expected to cause injury or death to persons or damage or
  destruction to, \emph{or neutralization of},
  objects.\footnote{\cite{revisedRule30}}
\end{quote}
\doublespacing

This amendment should not suggest that any disruption of service be regarded as
a cyber attack, but needs to allow for protection in cases like the attack on
Estonia, where there was a great effect on the nation's ability to operate
effectively, yet without causing any physical damage. Narrowing the
threshold that constitutes an armed attack remains an open issue.

Sections of the Tallinn Manual, and even this amended rule, could be seen as lacking some
detail or specifics. For example, when does an operation cross the threshold from an
inconvenience to an attack? Is the mere installation of malware a violation of
sovereignty? When is a State sufficiently involved with the creation or
distribution of an attack to warrant responsibility?
Much of this lack of detail in the Tallinn Manual can actually be regarded as
a strength through flexibility. Any corresponding kinetic issues are often
equally gray and require a case-by-case analysis. Cyberspace adds an additional layer of
complexity and rapid advancement. Making a set of rules as specific
and technically detailed as possible has the significant drawback of failing to
keep up with innovation and potentially allowing for legal loopholes. The
general nature of the laws allow for interpretation and for continual
operation in the spirit of those laws rather than the literal letter.

Potential adoption of the Tallinn Manual raises more than only technical
questions. Some of the greatest hurdles to a universal adoption of the Tallinn
Manual are political. As a NATO sponsored project, the translation of existing law to
the realm of a global cyberspace may be accompanied by a Western bias. Further, those
States typically at odds with NATO will be unlikely to consider a set of rules
compiled solely by NATO.

A subset of the currently undecided issues (whether neutral states can be held
accountable for only negligence, whether installing malware is itself a
violation of sovereignty, etc.) on which the Experts were divided could reasonably be more
definitive and solidified with time and
applied as general rules. Those States outside NATO, particularly those often in
disagreement, should be encouraged to participate in solidifying and defining
the rule set. This will help the Tallinn Manual to 1) be free of NATO bias and
2) be supported by all States, not just those who are members of NATO.

Once the evolution of the Tallinn Manual is made a global participation it will
be more likely to garner universal backing. When it does and appropriate
amendments are made, the current members of
the United Nations should be encouraged to ratify a treaty adopting adherence to
the rules set forth within the Tallinn Manual.

Any agreement should adopt the Tallinn Manual as a starting point for ruling on
international cyber conflict and be used as a basis to continually evolve relevant
international law. Given the rapid growth of information technology, it
should be a highest priority to establish an official means to make lasting and binding amendments
to the Tallinn Manual (or treaty), as deemed necessary by the Security Council and
members of the United Nations.

The United Nations already has an established means to judge violations of
established law, and the same framework should be used for cyberspace
agreements. Violation of the Tallinn Manual should be ruled on by the United Nations
Security Council, at which point any response, up to and including a use of force under
Article 51, is a potential outcome. This method of enforcement is also suggested by the Tallinn
Manual itself.\footnote{\cite{tallinnManual}, 66. See rule 18.}

\section{Where Adoption Brings the International Community}
The solution is neither complete nor final, but as with all law, we must
set up a solid base with which to work and progress over time. The
established international law made preparation for the age of the Internet
possible, and in the same way, the Tallinn Manual sets the stage for the
tools we need in decades to come, as the world grows more
reliant on the Internet.

The ability of a cyber operation to be classified as an armed attack, and by
extension the possibility of responding with military force under Article 51, is
quite significant and can lead to a severe escalation during times of
conflict. This is appropriate, though, as the world's critical infrastructure
becomes increasingly online and cyber weapons increase in ability and
complexity constantly. It is an antiquated notion that military arms must
explode, shoot, or even directly cause bloodshed. It is important that States
are granted the means to protect their cyberspace.

When asked about threshold of attacks and the significant damage that can be
inflicted through cyberspace, Schmitt implies he believes the standards for
classifying armed attacks in cyberspace will also evolve in coming years:

\singlespacing
\begin{quote}
  I anticipate that we'll see a lot of thresholds coming down that will
  allow States to respond more vibrantly to cyber attacks that  might not be
  possible under the law as we found it.\footnote{\cite{areThereIntlRules}.
    See paragraph 23.}
\end{quote}
\doublespacing

Overall, the conclusions of the Tallinn Manual feel just and reasonable.
Many of the issues not directly resolved are complex and likely require
analysis on a case-by-case basis over a number of years, all the while the laws will
be consistently developed and solidified. Achieving concurrence on acceptable
cyber acts will likely prove challenging, but it is a necessary aspect as the
cyber space continues to evolve in ways not yet considered.

Despite any unintentional bias introduced by the Experts' NATO background, their
effort to apply already agreed upon international law provides confidence in the
international arena that the international community, collectively, is
continuing to act in cyberspace in the spirit of previously established and
accepted laws.

Another notable, though indirect, potential effect of an established treaty
based on the rules of the Tallinn Manual is the preservation of an
open Internet within and among sovereign States. As the States of the world
realize the potential damage inflicted by attacks launched via cyberspace they are
exploring measures to regulate and police domestic networks, often at the
expense of some traditional Internet
freedoms.\footnote{\cite{declarationCyberIndependence}. The Electronic Frontier
  Foundation requested the sovereignty of cyberspace be respected
  by the world's governments. The demands of State infrastructure on the
  Internet have changed this landscape, but it may still be possible to preserve
  these ideals.}
For example, while the United States hasn't passed significant regulation on
networks and software to defend against cyber attacks, there are many arguments
to establish such proposals. Richard Clarke, a leading adviser on cyber
security in the United States political space, is a proponent for greater
regulations and had this to say of regulating the private sector:

\singlespacing
\begin{quote}
... industry only responds when you threaten regulation. If industry doesn't
respond (to the threat), you have to follow through.\footnote{\cite{clarkeRegulation}}
\end{quote}
\doublespacing

The openness of the Internet makes it a space unlike any other, and the
immediate censorship performed by totalitarian States proves it is a strong tool
for protecting human rights.\footnote{\cite{turkeyCensorship}. Turkey's Prime
  Minister censored Twitter and blocked Google's DNS after news of a potential
  scandal began spreading online.} So, while an indirect effect, the potential
preservation of a healthy, open cyberspace should not be understated, and
agreeing to the rules of the Tallinn Manual could help avoid heavy-handed
regulations.

\section{Conclusion}
The Tallinn Manual offers help in an area that the international community
needs in coming decades. There have been multiple recent occasions where cyber
operations have wreaked havoc on States' infrastructure, and the potential
consequences are likely to escalate as the world becomes more
connected. The risks are heightened by the certainty that offensive measures are
continuously advancing. It is crucial that States start taking the reasonable
measures to protect themselves and their citizens.

An application of the Tallinn Manual to the events surrounding the Stuxnet Worm
demonstrated that the work of Schmitt and the Experts have come a great way in
correlating existing law to operations in cyberspace. The Tallinn Manual is not
without technical and political hurdles, but it is ultimately worthy of serving
as a basis in establishing the first international laws for cyberspace.
The collective international community has an opportunity to
collaborate to work through the issues presented, ultimately offering a safer
and more fruitful cyberspace for all to utilize.

An agreement adhering to the rules of the Tallinn Manual would not only
help provide protection to the infrastructure and
sovereignty of participating States, it could encourage a continued free and
open Internet. The international community can look forward to a healthy,
global cyberspace with help from the Tallinn Manual and continued discussions on
permissible and responsible cyber operations and online stewardship.

Since the Tallinn Manual is inspired by already accepted international laws, it
can confidently be proposed as the beginning of a path forward. The Tallinn
Manual provides a strong foundation of which the international community should take
advantage.

\newpage
\singlespacing
\nocite{*}
\printbibliography

\end{document}
