% LaTeX for Steve Jarvis' CS 591

% 12pt required for project
\documentclass[12pt]{scrreprt}

% Now the actual document
\newcommand{\hwtitle}{Applicability and Adoption of the Tallinn Manual}
\newcommand{\hwsubtitle}{Considering Legality and Repercussions of the Stuxnet Worm}

\author{Steve Jarvis}
\title{ \hwtitle{} }
\subtitle{ \hwsubtitle{} }
\date{\today}

% Doc formatting, double space with 1.25 sides and 1'' top/bottom
\usepackage{setspace}
\doublespacing
\usepackage[margin=1.25in,top=1in,bottom=1in]{geometry}
\usepackage{enumerate}

% Page header
\usepackage{fancyhdr} \pagestyle{fancy} \lhead{Steve Jarvis} \rhead{\hwtitle{}}

% Bibliography
\usepackage[utf8]{inputenc}
\usepackage{biblatex-chicago}
\addbibresource{tallinn-and-stuxnet.bib}

\begin{document}
\maketitle

% body
\section*{Introduction}
The battlefield of the world is changing. For better or worse, militaries can
largely trade in their guns for laptops and their soldiers for hackers. The
nations of the world are connecting critical services and infrastructure to
cyberspace, and with minimal technical or legal protections in place. The
offensive capabilities and risks are continually increasing, so too must the
protections.

In the wake of the Second World War, the international community recognized the
necessity of establishing agreements and rules for international conflict.
The United Nations was organized in 1942\footnote{\cite{unitedNationsHistory}}
and the Geneva Conventions in 1949.\footnote{\cite{genevaConventionsOrigins}}
These organizations founded treaties that defined international laws for entering
into conflict (\textit{jus ad bellum}) and behavior during conflict
(\textit{jus in bello}), respectively.

The agreements by these groups were established a full half century before the
public internet was in its infancy, and as such, considered only kinetic
warfare. The recent growing prevalence of cyber operations is pushing the
boundaries of what might be considered harmful or forceful behavior, and the
international community has little means to deal with such events. Cyber
operations rarely have a direct or obvious kinetic
counterpart for comparison, and without established guidelines, states
continue to push the boundaries. Cyber operations are limited by technical
capabilities rather than legal or ethical standards.

The Tallinn Manual aims to fill the void of defined international law
when it comes to cyber
activity by translating the established kinetic laws to this modern
battlefield.\footnote{\cite{tallinnManual}, page 18, para 2.}

The goal of this report is to examine whether the Tallinn Manual is
suitable for adoption in the international community. To make this
determination, its applicability to real life actions, particularly the Stuxnet
worm, will be considered, as well as the Tallinn Manual's political hurdles. If
successful, the Tallinn Manual's adoption could help to ensure the safety and
security of all states in coming years.

\section*{Our Current Situation}
Today, 193 of 196 countries of the world are members of the United Nations.
%TODO have 193 agreed to the charter??
These countries enjoy agreements protecting their sovereignty and freedom from
unsolicited acts of physical aggression. These laws do not explicitly exclude
cyber acts of aggression, but the nature of international law is permissive--if
an act isn't explicitly stated illegal, then it's legal.\footnote{\cite{TODO see
citation 14 of tallinn}}

Differing interpretations of how established laws regulate cyber activity
results in dissension among states. There are no established answers on what
constitutes illegal activity, whether it's possible to exert a use of force or
launch an armed attack online, or what repercussions would be appropriate if
it is possible, to name a few.\footnote{\cite{cyberAttacksYale}, pages 431-435.}

With no consensus on what constitutes permissible cyber activity, there can be
no agreement on suitable international laws or treaties to govern behavior. By
extension, there are also no means to reprimand states or enforce acceptable
behavior in cyberspace.

\section*{Why This Needs to Change}
The worlds situation is only growing more dire as states grow increasingly more
reliant on cyberspace. The worlds' critical infrastructure, including power
grids, water treatment plants, financial institutions and health care equipment,
is growing increasingly more connected reliant on the Internet. Citizens and
governments rely on these services to maintain a healthy and safe standard of
living. This infrastructure, and by extension the cyberspace that
enables its proper operation, has a direct effect on a state's sovereignty and
deserves the same protections to which the physical world is privy.

The new realm of cyberspace provides a battleground the states of the world are
just beginning to explore. New attacks are constantly developed and far outpace
states' abilities to defend against them. The international community needs to
establish rules protecting states' cyberspace.

\section*{The Tallinn Manual}
The Tallinn Manual was an effort to fill the void in international law when it
comes to cyberspace. It was the effort of an international group of legal experts,
upon encouragement from the North Atlantic Treaty Organization (NATO), to
correlate existing international laws on conflict to
cyberspace.\footnote{\cite{tallinnManual}, page 16, para 1} The intent was to
give meaning to the laws established by the United Nations and Geneva
Conventions in cyberspace.

Over a period of three years, legal experts from many NATO states discussed the
complications of cyber operations and how to apply the established laws to this
new dimension with a clarity unachieved by any other attempts. The result is a
set of 25 Rules and many additional thoughts, issues and sticking points in the
form of commentary throughout the document.

The effort concluded in 2012 and continues to stand alone in its thoroughness
considering a modern world and outpaced laws.

\section*{The Stuxnet Worm}
Stuxnet was a revolutionary cyber attack in 2010 that sabotaged the programmable
logic controllers (PLCs) at Iran's Natanz nuclear enrichment facility. The
intent was to delay Iran's potential development of
nuclear weapons, and it did this by causing the centrifuges that spin uranium
to spin out of control and ultimately destroy
themselves.\footnote{\cite{symantecStuxnetOverview}} \footnote{\cite{stuxnetGoalResult}}

Despite the destructive chaos actually happening in the system, all the while
the hardware would deliver positive reports to the operators, assuring them that
operation was going safely and as planned, so the human operators capable of
intervening had no indication there was an issue, much less one so dire.

To consider Stuxnet an act of aggression by another nation, though, it has to be
known that it was in fact the product of a nation and not a private group of
citizens or hacktivists.\footnote{\cite{TODO}} Stuxnet's level of technical
difficulty was uncontested by any cyber attack
prior (and maybe since). Stuxnet was not only extremely complex, it was also
highly expensive.\footnote{\cite{stuxnetCostComplexity}} This evidence supports
the theory that the attack was orchestrated by a nation-state, which qualifies
it as a politically motivated operation. More recently, more concrete evidence
has surfaced that it was in fact executed by a nation-state, particularly a
collaboration between the United States and Israel.\footnote{\cite{TODO http://abcnews.go.com/blogs/headlines/2013/07/edward-snowden-u-s-israel-co-wrote-cyber-super-weapon-stuxnet/}}

Stuxnet exists amidst many controversial cyber operations over the past decade,
yet it still deserves the focus when considering cyber operations and
international law on armed conflict because it offers, arguably, the
most likely candidate of a cyber-only
operation to be considered a use of force or an armed attack. Prime alternative
operations include Flame Malware\footnote{\cite{flameForbes}}
and Russia's cyber attack on Georgia.\footnote{\cite{russiaAttackGeorgiaAC}}
Flame was used for only espionage, and that is generally considered legal under
international law.\footnote{\cite{tallinnManual}, rule 6, comment 4.} Russia's
cyber attack on Georgia in 2008, conversely, was used to further a kinetic
attack and falls clearly into the category of an armed
attack.\footnote{\cite{tallinnManual}, page 94, para 6
  and \cite{tallinnManual}, rule 20.} These operations raise many interesting
questions of their own, but the overall conclusions are unsurprising and generally
uncontroversial.

The enrichment facility at Natanz was not connected to the Internet. Despite
these precautions taken against cyber attack, it proved to still be
vulnerable. Stuxnet is believed to have been initially introduced to the target
environment by an infected USB drive.\footnote{\cite{stuxnetMicroscope}. See
  page 8.} En route to its final target, though, the virus crossed an unknown
number of state boundaries.

The damage was intended to stifle development of military operations by damaging
Iran's ability to enrich uranium.
The precise effects are unknown, as Iran refused to comment on whether it
suffered damage from Stuxnet. However, it's
widely held that their nuclear program did suffer significant setbacks, on the
order of what could have been accomplished with a kinetic
attack.\footnote{\cite{stuxnetIsis}}

As is with many worms, Stuxnet proved difficult to control. It leaked into the
wild and infected an untold number of civilian computers in more than a dozen
countries.\footnote{\cite{stuxnetMicroscope}. See page 15.} Eugene Kapersky, of
the Russian security company Kapersky Labs, claimed the virus even infected the
Russian nuclear facilities. He is not alone in these claims, either, with as
many as 15 other industrial complexes around the world believed to have been
infected.\footnote{\cite{stuxnetPlants}}

The distinction between civilian and nuclear or industrial facility infection
is significant, not only due to concerns
regarding civilian targeting, but because of the nature of Stuxnet itself.
Stuxnet was designed to recognize the hardware on which it was running, and if
it wasn't on specific Siemen's PLCs it would sit
dormant.\footnote{\cite{stuxnetPlants}}
Infected civilians would not only be unaware, they would be at no risk of danger
whatsoever. The infected industrial facilities, however, would likely suffer
grave damage, just as Iran's facility is believed to have experienced at
Natanz.\footnote{\cite{stuxnetIsis}}

\section*{Cyber Issues Where Traditional Law is Lacking}
Cyber operations like Stuxnet are unlike any with which the world has yet dealt.
The execution of these attacks raise many questions where guidance from a
suitable guide is desired to determine which laws, if any, were broken, how
severely they were broken, and what response might be warranted. There are a few
issues which warrant specific focus:

\subsection*{Use of Force}
At the most basic, it is not clear whether Stuxnet was even illegal under
international law. To determine its legality there would need to be consensus
that it did violate a law or laws already in place. One of the foremost laws
would be that prohibiting a use of force, set forth by Article 2(4) of the UN
Charter. It states:

\begin{quote}
  All Members shall refrain in their international relations from the threat or
  use of force against the territorial integrity or political independence of
  any state, or in any other manner inconsistent with the Purposes of the United
  Nations.\footnote{\cite{tallinnManual}. See Rule 11.}
\end{quote}

Stuxnet's classification as a use of force is not clear, however. Though the
distinction is not defined in the Charter, a traditional use of force involves
military force and violence, whereas Stuxnet, as most cyber attacks, did not
even include a physical presence.

\subsection*{Armed Attack}
Some states adopt a generally encompassing definition of what constitutes
illegal force, while others operate with much narrower
classifications.\footnote{\cite{tallinnManual}, page 17, para 3.} Supposing the
international community accepts the view that an armed attack always coincides a
use of force this issue is the same as the last, but this is unlikely. How might
an armed cyber attack be quantified?

Some popular definitions of an armed attack require it to be violent, political,
and (TODO this definition was in our reading one week). There have been no
cyber-only attacks to date to fit this description,
so is it even possible to launch an armed attack online?

\subsection*{Discriminate Attack}
The Stuxnet worm leaked into the wild and infected civilian and government
devices. A failure to control weapons in kinetic warfare is outlawed as
indiscriminate, but it's not certain the same applies to malware, in which there
is little danger of harm relative to illegal kinetic attacks, such as chemical
warfare.\footnote{\cite{genevaArticle51}} The civilian infection was not
necessarily harmful, and does mere infection equate to an attack in cyberspace?
Whether it was harmful or not, are there expectations for special consideration
of civilians in cyberspace?\footnote{\cite{TODO}}

The stakes are raised in the case of infected industry systems, however. Infection at
locations operating the targeted Siemens machines are subject to the same risk
and harm as the intended target, Iran.

\subsection*{Accountability}
It seems obvious that, if decided to be in violation of any laws, the sponsors
of the malware and its authors could be held responsible. There may be
additional participants to consider, though. In kinetic warfare,
belligerents cannot use neutral territory for transport of
munitions, and it is a duty of the neutral state to ensure that is the
case.\footnote{\cite{hagueConventionV5}. See Articles 2 and 5.} Is it legal
to move malware through the infrastructure of neutral states, and can those
states be held responsible for allowing it?
% Edgar suggested Hague is an outdated concept, but the Tallinn Manual itself
% quotes it (double check that)?

\subsection*{Retaliation, Defense and Rebuke}
Equally important is the issue of enforcement and reprimanding states who are
deemed in violation of any accepted law. The UN Charter allows for a use of
force as self defense in Article 51, but could a military response to
a computer hack be reasonable?\footnote{\cite{unCharter}. See Article 51.}

If Stuxnet or similar attacks are decided to fall short of a qualification as a
use of force can there be any retaliation taken? Particularly, what if the
victim state doesn't have the means to defend itself, yet the United Nations
doesn't deem the operation a use of force? This was the case in the 2008 attacks
Russia launched on Estonia, who suffered a completely debilitated Internet for
two weeks and received no aid after a request to the United Nations Security
Council.\footnote{\cite{TODO}} Would an external state with
sufficient capability have the United Nation's blessing to offer protection?

\section*{Applying the Tallinn Manual to the Stuxnet Worm}
% Determine the applicability of Tallinn Manual to Stuxnet. Does it work?
Now that there have been sufficient, specific issues raised with the
applicability of existing law to cyberspace, we turn to the Tallinn Manual to
help. We'd like to determine whether the Tallinn Manual is an effective tool in
practice by finding whether it proves useful when evaluating Stuxnet.

Regarding legality of Stuxnet, the Tallinn Manual defines a cyber operation to
be a use of force when the
effects of the operation ``are comparable to non-cyber operations rising to the
level of a use of force.''\footnote{\cite{tallinnManual}} It goes on to suggest
a detailed rating scale for cyber operations in the comments, including such
factors as severity, immediacy, directness, invasiveness and military character,
most of which Stuxnet clearly violates.\footnote{\cite{tallinnManual}. See page
  51.} The group of experts unanimously concluded Stuxnet to be a use of force,
and therefore in violation of Article 2(4) of the UN
Charter.\footnote{\cite{stuxnetForceWired2013}} Stuxnet also directly violates
at least Rule 1 of the Tallinn Manual.\footnote{\cite{tallinnManual}. See Rule 1.}

The manual goes on to say
that the experts were unanimous in determining cyber operations alone might
qualify as armed conflict.\footnote{\cite{tallinnManual}. See Rule 22, comment
  15.} Tallinn Manual states:

      \begin{quote}
        An armed conflict exists whenever there are here are hostilities, which
        may include or be limited to cyber operations, occurring between two or
        more States.\footnote{\cite{tallinnManual}. See Rule 22.}
      \end{quote}

This finding alone is significant, because it enables retaliation under Article 51
to a cyber attack within the realm of possibility. This directly contradicts
other interpretations of the Charter that reserve an armed attack for armed
military operation.\footnote{\cite{internationalLawForCyber}}

Tallinn Manual also comments directly that the experts were divided on whether
Stuxnet qualified as an ``armed attack''.\footnote{\cite{tallinnManual}. See Rule 22, comments 14
  and 15.} Some experts felt it was justified as anticipatory
self-defense.\footnote{\cite{tallinnManual} See Rule 13, comment 13.}

Indiscriminate methods of cyber war are
prohibited.\footnote{\cite{tallinnManual}. See Rule 43.} The infection of
civilians' computers (might\footnote{\cite{tallinnManual}. See Rule 1,
  comment 6.}) not violate this, because despite infection, the
attack was performed only on specific Siemens systems (not
civilian).\footnote{\cite{stuxnetGoalResult}} That is assuming, of course, the
mere installation of malware does not amount to an attack. It's not clear how
the Tallinn Manual would categorize such a thing.

If Kapersky's claims\footnote{\cite{stuxnetLeakKapersky}} that Stuxnet later
infected Russian nuclear sites are true, however, Stuxnet did violate this rule,
since Stuxnet was decidedly a cyber attack and this would prove it was not
limited in its effects as required by the law of armed
conflict.\footnote{\cite{genevaArticle51}}

Issues on which the Tallinn Manual was not conclusive. What self-defense is
legal when it's unclear whether an operation could be classified as an
``armed attack'', or when the Security Council has not yet reached an agreement?

As mentioned, Tallinn Manual is clear that cyber operations alone could amount to an armed
attack,\footnote{\cite{tallinnManual}, rule 22.} but provide little means to
determine whether a specific operation crosses that threshold,
or how an armed attack relates to a use of force (though that is a
standing problem with kinetic warfare, just the same). Schmitt says
these specifics were intentionally not included in the
Tallinn Manual.\footnote{\cite{areThereIntlRules}}

If we can assume an armed attack accompanies any use of force and
warrants a response under Article 51 (similar to the apparent view of
the United States (US)\footnote{\cite{cyberAttacksYale}.
  See page 433, para 4.}), then the Tallinn Manual provides a
fairly complete template to determine whether an operation
qualifies.\footnote{\cite{tallinnManual}. See rule 11, esp. comment 9.}

Schmitt claimed that:
\begin{quote}
  [...]the majority said that an attack, in the law of war, means
  you physically harm someone, you break something, you cause physical
  damage or you interfere in the functionality of an object such that it
  needs to be actually repaired.\footnote{\cite{areThereIntlRules},
    para 4.}
\end{quote}

This seems to qualify Stuxnet as an armed attack, yet other experts
still disagreed it crossed the undefined threshold. There was little
physical damage done, compared traditional, kinetic
force.\footnote{\cite{stuxnetArmedAttackRegister}}
There are also issues with immediacy that were not matters in kinetic
warfare. Iran didn't know it was being attacked until the threat was
over, and with no imminent or current attack there is no justification
for an armed response under Article 51.\footnote{\cite{tallinnManual}. See rule 15.}

Who can be held responsible in a cyber attack? Iran obviously had jurisdiction
in Stuxnet.\footnote{\cite{tallinnManual}. See Rule 2} Attribution is still a
significant challenge though.

What of neutral states whose infrastructure was used to transfer
Stuxnet? Laws on kinetic warfare imply they can be held accountable for
transferring munitions. The Hague Conventions
itself limits a neutral state's obligation to restrict the combatant use
of telephone or wireless cables.\footnote{\cite{hagueConventionV5}.
  See article 8.} Tallinn Manual states it must not
\emph{knowingly} allow conflicting parties to use its
infrastructure.\footnote{\cite{tallinnManual}. See rule 93.} The experts
were split on whether that permits neutral states to passively allow
such transmission.\footnote{\cite{tallinnManual} See rule 93, comments 5 and 6.}

\section*{Adopting the Tallinn Manual}
% Proposed solution. So is it's effective, what should we do about it?
Proposed solution for resolving international cyber conflict.
The Tallinn Manual does an outstanding job of applying existing
international law to the age of the internet, albeit occasionally lacking some
detail. This indecisiveness can also be seen as a strength. Much of the gray
area is equally undecided in kinetic
conflict, terribly complex and likely to evolve at a rapid pace.

The Tallinn Manual should be adopted as a starting point for ruling on
international cyber conflict and used to evolve international law, just as
the US Constitution served as a starting point for US law and was molded
by hundreds of years of precedence.

Adoption of the Tallinn Manual raises more than only technical questions.
Some of the greatest hurdles to adopting the Tallinn Manual are political.
As a NATO sponsored project, the translation of existing law to
the realm of the internet may be accompanied by a Western bias. Further, those
states usually at odds with NATO will be unlikely to consider a set of rules
built solely by NATO.

Some of the currently conflicting issues (responsibility of neutral
states, whether installing malware is itself a violation of sovereignty,
etc) on which the experts were divided could be solidified with time and
applied as general rules. Those states outside NATO, particularly those often in
disagreement, should be encouraged to participate in solidifying and defining
the rule set. This will help the Tallinn Manual to 1) be free of NATO bias and
2) be supported by all states, not just those who are members of NATO.

Once the evolution of the Tallinn Manual is made a global participation, the
current members of the UN should be encouraged to sign a treaty
adopting adherence to the rules it defines. This is the ultimate goal.

Violation of the Tallinn Manual should be ruled on by the UN Security
Council, at which point any retaliation, up to and including that under
Article 51, is a potential outcome. This is also suggested in the Tallinn
Manual itself.\footnote{\cite{tallinnManual}, rule 18.}

There should be an official means to make lasting and binding amendments
to the Manual (or treaty), as deemed necessary by the Security Council and
members of the UN.

The efforts of the Tallinn Manual will give us confidence in the
international arena that we, collectively, are continuing to act in cyber
space in the spirit of already established and accepted international laws.

\section*{Where Adoption Brings International Cyber Space}
Analysis of the solution.
The solution is neither complete nor final, but as with all law, we must
set up a solid base with which to work and progress over time. The
established international law made preparation for the age of the internet
possible, and in the same way, the Tallinn Manual sets the stage for the
tools we need in decades to come, as the world grows increasingly more
reliant on the Internet.
Schmitt implied he believes the standards for classifying armed attacks
in cyberspace will evolve in coming years:

\begin{quote}
  I anticipate that we'll see a lot of thresholds coming down that will
  allow states to respond more vibrantly to cyber attacks that  might not be
  possible under the law as we found it.\footnote{\cite{areThereIntlRules},
    para 23.}
\end{quote}

The ability of a cyber operation to be classified as an armed attack, and by
extension the possibility of responding with military force under Article 51, is
quite significant and can lead to a severe escalation during times of
conflict. This is appropriate, though, as the world's critical infrastructure
grows increasingly more online and cyber weapons increase in ability and
complexity constantly. It is an antiquated notion that military ``arms'' must
explode, shoot, or even directly cause bloodshed.

The conclusions of the Tallinn Manual feel reasonable.
Many of the issues not directly resolved are complex and likely require
analysis on a case-by-case basis. The issues that were directly resolved
are not only founded on accepted international law, but feel just when
applied to the cyber world.

Achieving concurrence on acceptable cyber acts will likely prove
challenging, but it is a necessary feature as the cyber space continues to
evolve in ways not yet considered.

\section*{Conclusion}
Conclusion
The Tallinn Manual offers help in an area which the international community much
needs it in the coming years. It should be taken advantage of.

The Tallinn Manual is worthy of official acceptance in founding international
law and standards. The international community can look forward to a healthy,
global internet with help from the Tallinn Manual and subsequent discussions on
permissible and responsible online operations and stewardship.
It is the confidence to move forward.

\newpage
\nocite{*}
\printbibliography

\end{document}
